### 2017-12-20 chapters/paper1/article.tex
Fix labels and reference in the papers

% \begin{center}
Gabriel Balaban$^{1,2,5}$\footnotemark[\value{footnote}],
Henrik Finsberg$^{1,2,5}$\footnotemark,
Hans Henrik Odland$^{4,7}$,
M. E. Rognes$^{1,3}$,
Stian Ross$^{4,4}$,
Joakim Sundnes$^{1,2,5}$, and
Samuel Wall$^{1,5,6}$

@@

The rest of this paper is organized as follows: In
Section~\ref{sec:methods} we present a mathematical model that
accounts for the three main drivers of ventricular mechanics; blood
pressure, tissue elasticity and muscle contraction.  
We also describe clinical measurements of a patient suffering from dyssynchrony, and our data assimilation procedure
for fitting the model to these measurements.
Numerical results are presented in Section~\ref{sec:num_results}, and discussed in
Section~\ref{sec:discussion}. Finally, we provide some concluding
remarks in Section~\ref{sec:conclusion}.

\section{Materials and Methods}
\label{sec:methods}

@@

\label{sec:wall_motion}
In order to estimate the position of the myocardial walls through
the cardiac cycle we adopt a continuum mechanics description of
cardiac wall motion. In this description we consider a fixed left
ventricular reference geometry $\Omega$, with endocardial boundary
$\lvendo$, and basal boundary $\basebound$.

@@

Assuming that the cardiac walls are in equilibrium,
it is possible to determine the value of $\uvec$ from the principle of
virtual work
\begin{equation}
  \label{eq:work-balance}
  \delta W(\uvec) = 0,
\end{equation}
which states that the virtual work, $\delta W(\uvec)$, of all forces
applied to a mechanical system vanishes in equilibrium. For our ventricular wall motion
model, the virtual work $\delta W(\uvec)$, is given by
\begin{equation}
\begin{aligned}
\label{eq:workdef}
 \delta W(\uvec, p) &= \int_{\Omega} \FPK : \Grad \ \du \ dV 
		  + \int_{\Omega} (J - 1) \dpre + p J \F^{-T} : \Grad \ \du \ dV  \\
		 &+ \plv \int_{\lvendo} J \F^{-T} \Nvec \cdotp \du \ dS 
		  +  \int_{\lvbase}  k \uvec \cdotp \du \ dS.
 \end{aligned}
\end{equation}
Here we have introduced the hydrostatic pressure $p$ in order to enforce the 
incompressibility constraint $J =1$, with $J = \det \F = \det  \left(\Grad \ \uvec + \mathbf{I}\right)$, 
and $\mathbf{I}$ being the second order identity tensor. 
Furthermore, $\Nvec$ denotes the unit outward normal vector, $k$ 
the constant of a spring that we introduce at the basal boundary, and $\plv$ the intra-ventricular 
blood pressure. The virtual variables $\du$ and $\dpre$ are test functions whose values
are arbitrary when the system \eqref{eq:work-balance} is in mechanical
equilibrium.

In order to anchor the computational geometry, we fix $\uvec$ in the longitudinal direction 
at the base by using a Dirichlet boundary condition. At the epicardial boundary normal
forces are set to 0, and so there is no term for this boundary in
\eqref{eq:workdef}.

The internal stresses of our model are given by $\FPK$, the first Piola-Kirchhoff tensor,
which can be calculated as a derivative of a strain energy functional
in the case of a hyperelastic material.
In our model we employ a reduced version~\cite{Krishnamurthy2013, gjerald2015patient,
  asner2015estimation, hadjicharalambous2015analysis} of the
Holzapfel-Ogden strain energy law~\cite{Holzapfel2009}, 
\begin{equation}
\label{eq:hoa}
 \psi(\C) = \frac{a}{2 b} \left( e^{ b (I_1(\C)
 - 3)}  -1 \right)
 + \frac{a_f}{2b_f} \left( e^{ b_f (I_{4f}(\C)
 - 1)_+^2} -1 \right),
\end{equation}
which gives the amount of strain energy, $\psi$, stored per unit
volume myocardium undergoing the strain $\C = \F^{T} \F$.
The notation
$(\cdotp)_{+}$ refers here to $\max\{\cdotp, 0\}$.  Furthermore
the mechanical invariants $I_1$ and $I_{4f}$ are defined as
\begin{equation}
I_1(\C) = \tr \C,  \quad \quad I_{4f} = \ef \cdotp \C \ef,
\end{equation}
with $\ef$ indicating the local myocardial fiber direction. The
material parameters $a, a_f, b, b_f$ are scalar quantities which influence the shape of
the stress-strain relationship, and can be adapted to personalize the
elastic properties of a myocardial tissue model to a specific patient.

The Lagrange multiplier formulation of incompressibility that we 
 employ enforces its constraint only weakly. This can cause
 convergence issues in the numerical solution of the work balance equation~\eqref{eq:work-balance}. We therefore
 eliminate volumetric strains from the energy function \eqref{eq:hoa}
 by a simple modification
 \begin{equation}
  \tilde{\psi}(\C) = \psi(J^{-\frac{2}{3}}\C).
 \label{eq:strain_energy}
 \end{equation}
 This modification has been shown to improve the robustness of
 Newton-Raphson methods applied to incompressible hyperelastic
 problems~[Figure 3C of \cite{land2015improving}].

In order to account for muscle contraction we apply the 
active strain framework~\cite{nardinocchi2007active}.  In this framework the amount
of muscle fiber shortening is specified by a field $\gamma$ via
a split of the deformation gradient
\begin{equation}
  \label{eq:fsplit}
  \F = \F_e \F_a(\gamma),
\end{equation}
where $\F_e$ is the elastic part and $\F_a(\gamma)$ the active
part of the deformation gradient. For the value of $\F_a(\gamma)$
we adopt a simple relation~\cite{gjerald2015patient, evangelista2011torsion} which satisfies
the incompressibility constraint by design and directly relates the
amount of active fiber shortening to the value of $\gamma$
\begin{equation}
  \F_a = (1 - \gamma) \ef \otimes \ef  + \frac{1}{\sqrt{1 - \gamma}} (\I- \ef \otimes \ef).
 \label{eq:active_strain}
\end{equation}
In the case $\gamma = 0$ there is no muscle shortening at all, and the
amount of shortening increases with increased $\gamma$ up to the
theoretical limit of $\gamma = 1$. Physiologically, $\gamma$ models the 
length change along muscle fibers neglecting elastic effects. This, together with the elastic
resistance gives the strength of the muscle contraction.

Muscle contraction is accounted for in terms of virtual work by
modifying the first Piola-Kirchhoff stress tensor, so
that the strain energy only depends on the elastic part of the
deformation
\begin{equation}
 \FPK = \frac{\partial \tilde{\psi}}{\partial \F} = \frac{\partial \tilde{\psi} (\C_e)}{\partial \F}
\label{eq:active_strain_energy}
\end{equation}
with $\C_e = \F_e^T\F_e$.

Given an amount of fiber shortening $\gamma$, the value of the elastic
 parameters $a, b, a_f, b_f$, the intraventricular blood pressure
 $\plv$ and the spring constant $k$, the myocardial wall displacement
 $\uvec$ and hydrostatic pressure $p$ can be obtained by solving the
 principle of virtual work \eqref{eq:work-balance}.

 \subsection{Clinical measurements}
\label{sec:clinical_measurements}
Clinical data were obtained at the Oslo
University Hospital in the context of the Impact study \cite{ImpactStudy2016}. Specifically,
we consider the case of an 82 year old man in NYHA functional 
class III systolic heart failure with coronary
artery disease, and left bundle branch block.
A left bundle branch block normally causes both electrical and mechanical 
dyssynchrony. In this case the ECG revealed a
QRS width of 140 ms and the echocardiographically derived ejection fraction of 30 \%.

@@

Valvular events were used to synchronize the pressure to the strain and
volume data. The timing of the observed valvular events in the images
were matched with the observed valvular events in the pressure
trace. In the pressure trace, aortic valve opening (AVO) was selected after the
steepest increase of the pressure ($\frac{dp}{dt}$ max ), and mitral valve
closure just before $\frac{dp}{dt}$ max . Aortic
valve closure (AVC) was chosen just before the pressure had its largest
decrease after AVO, and the mitral valve opening
before the pressure dropped down to baseline after AVC. 
A pressure-volume loop based on the synchronization is displayed in
Figure~\ref{fig:pv_loop_phases}.

@@

\subsection{Ventricular geometry generation}
\label{sec:compgeo}
The computational mechanics framework used for our wall motion model,
described in Section~\ref{sec:wall_motion}, requires a reference stress-free geometry from
which to define displacements. Such a geometry typically does not exist in-vivo due to the 
presence of blood pressure on the endocardial walls. Algorithms exist for calculating
stress free geometries given a loaded state \cite{bols2013computational, gee2010computational}.
However for the sake of simplicity we derive our reference geometry from an echocardiographic
image of the LV at the beginning of atrial systole,
as the pressure is near minimal at this point, and the ventricular myocardium
can be assumed to be relaxed.

@@

We employed Gmsh \cite{geuzaine2009gmsh} to create a 
linear tetrahedral volumetric mesh between the endocardial and epicardial surfaces. 
This mesh had 1262 elements, and is shown in Figure~\ref{fig:compgeo_mesh}.
Myocardial fiber orientations were assigned using a rule based method,
with a fiber helix angle of $40$ degrees on the 
endocardium rotated clockwise throughout the ventricular wall 
to $-50$ degrees on the epicardium \cite{bayer2012novel}. 
A streamline representation of the local myocardial fibers is displayed
in Figure \ref{fig:compgeo_fibers}.

Finally, the AHA-segments from the strain mesh were transferred onto the volumetric mesh.
This was accomplished by computing prolate spherical coordinates for
the barycenter of each tetrahedron,
and then assigning an AHA-zone to the tetrahedron 
based on the corresponding prolate spherical coordinate in the strain mesh. 
AHA-segments on the volumetric mesh are shown in Figure~\ref{fig:compgeo_regions}.

\begin{figure}[htbp]
\centering
\begin{subfigure}[t]{0.25\textwidth}
     \centering
     \raisebox{2mm}{\includegraphics[width=0.9\textwidth]{LV_systole_volume_trace}}
     \caption{\label{fig:ultrasound_image}}
\end{subfigure}
\qquad
\begin{subfigure}[t]{0.25\textwidth}
    \includegraphics[width=\textwidth, trim={15cm 4cm 21cm 3cm}, clip]{mesh}
    \caption{\label{fig:compgeo_mesh}}    
\end{subfigure}
\qquad
\begin{subfigure}[t]{0.25\textwidth}
    \includegraphics[width=\textwidth, trim={15cm 4cm 21cm 3cm}, clip]{fibers}
    \caption{\label{fig:compgeo_fibers}}
\end{subfigure}
\qquad
\begin{subfigure}[t]{0.25\textwidth}
    \includegraphics[width=\textwidth, trim={15cm 4cm 21cm 3cm}, clip]{sfun}
    \caption{\label{fig:compgeo_regions}}
\end{subfigure}
\begin{subfigure}[t]{0.35\textwidth}
    \includegraphics[width=\textwidth]{bullseye}
    \caption{\label{fig:bullseye}}
  \end{subfigure}
\caption{Ventricular geometry generation. Endo and epi-cardial surfaces 
are marked on 3-D ultrasound images. Figure (\ref{fig:ultrasound_image}) shows the endocardial 
marking for a 2-D slice of one such image. Next a computational geometry is generated
from epi and endo-cardial surfaces (\ref{fig:compgeo_mesh}), and rule based
fibers are assigned (\ref{fig:compgeo_fibers}). Finally AHA segments are assigned
to the geometry (\ref{fig:compgeo_regions}), according to the standardized scheme
(\ref{fig:bullseye}).}
\end{figure}

\subsection{Parameter Estimation}
\label{sec:paramest}
Now that we have a mathematical description of cardiac motion, along
with a personalized computational geometry and target data, we next turn to the
problem of personalizing the motion model via the estimation of the
elastic parameters and the fiber contraction. As dyssynchrony is a disease 
which primarily effects the contraction properties of the ventricle,
we focus our efforts on contraction modelling and employ a very simple
personalization of stiffness properties. That is only the parameter $a$
is optimized to fit the ventricular volumes, and the other elastic parameters
are kept fixed at the values ($a_f = 1.685, b = 9.726, b_f = 15.779$),
which were obtained from a bi-axial loading experiment 
[Table 1 row 3 of \cite{Holzapfel2009}].

@@

Muscle shortening is typically present in the ventricles
throughout systole and in early diastole until the muscles fully release
their contraction. During the phase of atrial systole we do not expect
muscle contraction in the ventricle, and so we set $\gamma = 0$ for
this phase. This allows us to estimate elastic properties
independently of contraction during atrial systole, and then estimate
contraction at each point in the rest of the cardiac cycle with the
material parameters fixed. In Figure~\ref{fig:pv_loop_phases} we show
the pressure-volume loop of the patient under consideration, and
highlight the phases where we estimate the contraction and elastic
parameters.

\begin{figure}[htbp]
\centering
    \includegraphics[width=0.8\textwidth]{patient_pv_loop}
\caption{Patient pressure-volume relationship for the left
  ventricle. Measurements in the blue solid line are used to estimate
  contraction, whereas measurements in the red dashed line are used
  to estimate elasticity. }
\label{fig:pv_loop_phases}
\end{figure}

\subsection{Definition of functionals}
\label{sec:data_func}
As described in Section~\ref{sec:clinical_measurements}, the data
available for our personalization of the wall motion model are
pressure, volume and strain measurements throughout the cardiac
cycle. The pressure measurements are included in the model as a boundary condition 
via the virtual work~\eqref{eq:work-balance},
and thus our data assimilation only needs to fit the model to
the volume and strain measurements. This requires that
we define a suitable set of functionals that quantify the
model-strain and model-volume mismatches. The personalization of the wall motion model can
then be achieved by optimizing the contraction and elastic parameters
in order to minimize the total model-data mismatch.

Let $i$ denote the index of an observed cavity volume $V^i$, or strain
$\varepsilon^i$, in the cardiac cycle. Furthermore let $j \in
\{1,..,17\}$ be the index of an AHA segment $\Omega_j$, and $k \in
\{c,r,l\}$ indicate a direction: circumferential, radial or
longitudinal, respectively. Given a measurement point $i$, we define
the model-strain mismatch
\begin{equation}
 \Istrain^i =  \sum_{j= 1}^{17} \sum_{k \in \{c,r,l\}}  \left( \varepsilon_{kj}^i -  \tilde{\varepsilon}_{kj}^i \right)^2,
\label{eq:strain_functional}
 \end{equation}
for model strain $\tilde{\varepsilon}_{kj}^i$ and measured strain
$\varepsilon_{kj}^i$. The speckle tracking software we use provides the measured strain $\varepsilon_{kj}^i$. 
This strain is regionally averaged and of Lagrangian type. In order to mimic this in our model we 
define the model strain as
\begin{equation}
\varepsilon_{kj} =  \frac{1}{|\Omega_j|}\int_{\Omega_j}  \mathbf{e}_k^T \nabla \uvec \, \mathbf{e}_k \ dx,
\label{eq:g_futhese nctional}
\end{equation}
where $\ek$ denotes a unit direction field and
$|\Omega_j|$ the volume of segment $j$.

Furthermore we also define the model-volume mismatch
\begin{equation}
  \Ivol^i  =  \left( \frac{V^i - \tilde{V}^i}{V^i} \right)^2,
\end{equation}
where the model volume is calculated by the formula
\begin{equation}
  \tilde{V}^i = -\frac{1}{3} \int_{\lvendo} (\Xvec + \uvec) \cdotp J \F^{-T} \Nvec  \ dS.
  \label{eq:vol_functional}
 \end{equation}
We note that this method of calculating the model volume depends upon
$(\Xvec + \uvec)\cdotp \Nvec = 0$ at the basal plane. These conditions hold in 
our model as the basal plane is defined with 0 longitudinal coordinate and 
longitudinal displacements are also set to 0 in this plane.

In order to have a single optimization target to describe the fit to data 
we combine the strain \eqref{eq:strain_functional} and volume
\eqref{eq:vol_functional} mismatches into one single functional
\begin{equation}
  \Idata^i(\alpha) = \alpha \Ivol^i + (1 - \alpha) \Istrain^i.
\label{eq:total_functional}
\end{equation}
Here the parameter $\alpha$ controls the relative emphasis of the parameter
estimation on volume or strain matching.

In our study we consider a high dimensional representation of $\gamma$ in order
to more accurately capture the details of a dyssynchronous contraction. However this can easily lead to
an over-parametrized problem in which many parameter combinations produce the same
functional values. In order to further constrain the optimization we introduce
a 1st order Tikohonov regularization functional
\begin{equation}
 \Ismooth^i(\lambda) = \lambda \| \nabla \gamma^i \|^2, 
 \label{eq:I_smooth}
\end{equation}
where $\|\cdotp \|$ represents the standard $L^2$ norm.
This functional discriminates between $\gamma$ parameter sets based on their smoothness.
The parameter $\lambda$ can be adjusted to control the size of the functional and hence
the relative emphasis on smoothing.

\subsection{Parameter estimation as an optimization problem}
\label{sec:param_est}
The elastic parameters of the reduced
Holzapfel-Ogden law \eqref{eq:hoa} represent the passive elastic
properties of the myocardium. We personalize these properties by
adjusting the parameter $a$ to match measured left ventricular volumes.
Mathematically this problem is formulated as
\begin{equation}
\begin{aligned}
\label{eq:opt_matparam}
& \underset{a}{\text{minimize}}
& &  \sum_{i = 1}^{\Nas} \Ivol^i \\
& \text{subject to}
& & \delta W(\plv^i, a) = 0 \quad \forall \,  i \ \in \{1,...N_{AS} \},
\end{aligned}
\end{equation}
where $\delta W$ is given by \eqref{eq:workdef}, and $\Nas =
3$ indicates the total number of measurements available in atrial
systole.

>>>>

% \begin{center}
Gabriel Balaban$^{1,2,5}$\footnotemark[\value{footnote}],
Henrik Finsberg$^{1,2,5}$\footnotemark,
Hans Henrik Odland$^{4,5,7}$,
M. E. Rognes$^{1,3}$,
Stian Ross$^{4,5}$,
Joakim Sundnes$^{1,2,5}$, and
Samuel Wall$^{1,5,6}$

@@

The rest of this paper is organized as follows: In
Section~\ref{paper1:sec:methods} we present a mathematical model that
accounts for the three main drivers of ventricular mechanics; blood
pressure, tissue elasticity and muscle contraction.  
We also describe clinical measurements of a patient suffering from dyssynchrony, and our data assimilation procedure
for fitting the model to these measurements.
Numerical results are presented in Section~\ref{paper1:sec:num_results}, and discussed in
Section~\ref{paper1:sec:discussion}. Finally, we provide some concluding
remarks in Section~\ref{paper1:sec:conclusion}.

\section{Materials and Methods}
\label{paper1:sec:methods}

@@

\label{paper1:sec:wall_motion}
In order to estimate the position of the myocardial walls through
the cardiac cycle we adopt a continuum mechanics description of
cardiac wall motion. In this description we consider a fixed left
ventricular reference geometry $\Omega$, with endocardial boundary
$\lvendo$, and basal boundary $\basebound$.

@@

Assuming that the cardiac walls are in equilibrium,
it is possible to determine the value of $\uvec$ from the principle of
virtual work
\begin{equation}
  \label{paper1:eq:work-balance}
  \delta W(\uvec) = 0,
\end{equation}
which states that the virtual work, $\delta W(\uvec)$, of all forces
applied to a mechanical system vanishes in equilibrium. For our ventricular wall motion
model, the virtual work $\delta W(\uvec)$, is given by
\begin{equation}
\begin{aligned}
\label{paper1:eq:workdef}
 \delta W(\uvec, p) &= \int_{\Omega} \FPK : \Grad \ \du \ dV 
		  + \int_{\Omega} (J - 1) \dpre + p J \F^{-T} : \Grad \ \du \ dV  \\
		 &+ \plv \int_{\lvendo} J \F^{-T} \Nvec \cdotp \du \ dS 
		  +  \int_{\lvbase}  k \uvec \cdotp \du \ dS.
 \end{aligned}
\end{equation}
Here we have introduced the hydrostatic pressure $p$ in order to enforce the 
incompressibility constraint $J =1$, with $J = \det \F = \det  \left(\Grad \ \uvec + \mathbf{I}\right)$, 
and $\mathbf{I}$ being the second order identity tensor. 
Furthermore, $\Nvec$ denotes the unit outward normal vector, $k$ 
the constant of a spring that we introduce at the basal boundary, and $\plv$ the intra-ventricular 
blood pressure. The virtual variables $\du$ and $\dpre$ are test functions whose values
are arbitrary when the system \eqref{paper1:eq:work-balance} is in mechanical
equilibrium.

In order to anchor the computational geometry, we fix $\uvec$ in the longitudinal direction 
at the base by using a Dirichlet boundary condition. At the epicardial boundary normal
forces are set to 0, and so there is no term for this boundary in
\eqref{paper1:eq:workdef}.

The internal stresses of our model are given by $\FPK$, the first Piola-Kirchhoff tensor,
which can be calculated as a derivative of a strain energy functional
in the case of a hyperelastic material.
In our model we employ a reduced version~\cite{Krishnamurthy2013, gjerald2015patient,
  asner2015estimation, hadjicharalambous2015analysis} of the
Holzapfel-Ogden strain energy law~\cite{Holzapfel2009}, 
\begin{equation}
\label{paper1:eq:hoa}
 \psi(\C) = \frac{a}{2 b} \left( e^{ b (I_1(\C)
 - 3)}  -1 \right)
 + \frac{a_f}{2b_f} \left( e^{ b_f (I_{4f}(\C)
 - 1)_+^2} -1 \right),
\end{equation}
which gives the amount of strain energy, $\psi$, stored per unit
volume myocardium undergoing the strain $\C = \F^{T} \F$.
The notation
$(\cdotp)_{+}$ refers here to $\max\{\cdotp, 0\}$.  Furthermore
the mechanical invariants $I_1$ and $I_{4f}$ are defined as
\begin{equation}
I_1(\C) = \tr \C,  \quad \quad I_{4f} = \ef \cdotp \C \ef,
\end{equation}
with $\ef$ indicating the local myocardial fiber direction. The
material parameters $a, a_f, b, b_f$ are scalar quantities which influence the shape of
the stress-strain relationship, and can be adapted to personalize the
elastic properties of a myocardial tissue model to a specific patient.

The Lagrange multiplier formulation of incompressibility that we 
 employ enforces its constraint only weakly. This can cause
 convergence issues in the numerical solution of the work balance equation~\eqref{paper1:eq:work-balance}. We therefore
 eliminate volumetric strains from the energy function \eqref{paper1:eq:hoa}
 by a simple modification
 \begin{equation}
  \tilde{\psi}(\C) = \psi(J^{-\frac{2}{3}}\C).
 \label{paper1:eq:strain_energy}
 \end{equation}
 This modification has been shown to improve the robustness of
 Newton-Raphson methods applied to incompressible hyperelastic
 problems~[Figure 3C of \cite{land2015improving}].

In order to account for muscle contraction we apply the 
active strain framework~\cite{nardinocchi2007active}.  In this framework the amount
of muscle fiber shortening is specified by a field $\gamma$ via
a split of the deformation gradient
\begin{equation}
  \label{paper1:eq:fsplit}
  \F = \F_e \F_a(\gamma),
\end{equation}
where $\F_e$ is the elastic part and $\F_a(\gamma)$ the active
part of the deformation gradient. For the value of $\F_a(\gamma)$
we adopt a simple relation~\cite{gjerald2015patient, evangelista2011torsion} which satisfies
the incompressibility constraint by design and directly relates the
amount of active fiber shortening to the value of $\gamma$
\begin{equation}
  \F_a = (1 - \gamma) \ef \otimes \ef  + \frac{1}{\sqrt{1 - \gamma}} (\I- \ef \otimes \ef).
 \label{paper1:eq:active_strain}
\end{equation}
In the case $\gamma = 0$ there is no muscle shortening at all, and the
amount of shortening increases with increased $\gamma$ up to the
theoretical limit of $\gamma = 1$. Physiologically, $\gamma$ models the 
length change along muscle fibers neglecting elastic effects. This, together with the elastic
resistance gives the strength of the muscle contraction.

Muscle contraction is accounted for in terms of virtual work by
modifying the first Piola-Kirchhoff stress tensor, so
that the strain energy only depends on the elastic part of the
deformation
\begin{equation}
 \FPK = \frac{\partial \tilde{\psi}}{\partial \F} = \frac{\partial \tilde{\psi} (\C_e)}{\partial \F}
\label{paper1:eq:active_strain_energy}
\end{equation}
with $\C_e = \F_e^T\F_e$.

Given an amount of fiber shortening $\gamma$, the value of the elastic
 parameters $a, b, a_f, b_f$, the intraventricular blood pressure
 $\plv$ and the spring constant $k$, the myocardial wall displacement
 $\uvec$ and hydrostatic pressure $p$ can be obtained by solving the
 principle of virtual work \eqref{paper1:eq:work-balance}.

 \subsection{Clinical measurements}
\label{paper1:sec:clinical_measurements}
Clinical data were obtained at the Oslo
University Hospital in the context of the Impact study \cite{ImpactStudy2016}. Specifically,
we consider the case of an 82 year old man in NYHA functional 
class III systolic heart failure with coronary
artery disease, and left bundle branch block.
A left bundle branch block normally causes both electrical and mechanical 
dyssynchrony. In this case the ECG revealed a
QRS width of 140 ms and the echocardiographically derived ejection fraction of 30 \%.

@@

Valvular events were used to synchronize the pressure to the strain and
volume data. The timing of the observed valvular events in the images
were matched with the observed valvular events in the pressure
trace. In the pressure trace, aortic valve opening (AVO) was selected after the
steepest increase of the pressure ($\frac{dp}{dt}$ max ), and mitral valve
closure just before $\frac{dp}{dt}$ max . Aortic
valve closure (AVC) was chosen just before the pressure had its largest
decrease after AVO, and the mitral valve opening
before the pressure dropped down to baseline after AVC. 
A pressure-volume loop based on the synchronization is displayed in
Figure~\ref{paper1:fig:pv_loop_phases}.

@@

\subsection{Ventricular geometry generation}
\label{paper1:sec:compgeo}
The computational mechanics framework used for our wall motion model,
described in Section~\ref{paper1:sec:wall_motion}, requires a reference stress-free geometry from
which to define displacements. Such a geometry typically does not exist in-vivo due to the 
presence of blood pressure on the endocardial walls. Algorithms exist for calculating
stress free geometries given a loaded state \cite{bols2013computational, gee2010computational}.
However for the sake of simplicity we derive our reference geometry from an echocardiographic
image of the LV at the beginning of atrial systole,
as the pressure is near minimal at this point, and the ventricular myocardium
can be assumed to be relaxed.

@@

We employed Gmsh \cite{geuzaine2009gmsh} to create a 
linear tetrahedral volumetric mesh between the endocardial and epicardial surfaces. 
This mesh had 1262 elements, and is shown in Figure~\ref{paper1:fig:compgeo_mesh}.
Myocardial fiber orientations were assigned using a rule based method,
with a fiber helix angle of $40$ degrees on the 
endocardium rotated clockwise throughout the ventricular wall 
to $-50$ degrees on the epicardium \cite{bayer2012novel}. 
A streamline representation of the local myocardial fibers is displayed
in Figure \ref{paper1:fig:compgeo_fibers}.

Finally, the AHA-segments from the strain mesh were transferred onto the volumetric mesh.
This was accomplished by computing prolate spherical coordinates for
the barycenter of each tetrahedron,
and then assigning an AHA-zone to the tetrahedron 
based on the corresponding prolate spherical coordinate in the strain mesh. 
AHA-segments on the volumetric mesh are shown in Figure~\ref{paper1:fig:compgeo_regions}.

\begin{figure}[htbp]
\centering
\begin{subfigure}[t]{0.25\textwidth}
     \centering
     \raisebox{2mm}{\includegraphics[width=0.9\textwidth]{LV_systole_volume_trace}}
     \caption{\label{paper1:fig:ultrasound_image}}
\end{subfigure}
\qquad
\begin{subfigure}[t]{0.25\textwidth}
    \includegraphics[width=\textwidth, trim={15cm 4cm 21cm 3cm}, clip]{mesh}
    \caption{\label{paper1:fig:compgeo_mesh}}    
\end{subfigure}
\qquad
\begin{subfigure}[t]{0.25\textwidth}
    \includegraphics[width=\textwidth, trim={15cm 4cm 21cm 3cm}, clip]{fibers}
    \caption{\label{paper1:fig:compgeo_fibers}}
\end{subfigure}
\qquad
\begin{subfigure}[t]{0.25\textwidth}
    \includegraphics[width=\textwidth, trim={15cm 4cm 21cm 3cm}, clip]{sfun}
    \caption{\label{paper1:fig:compgeo_regions}}
\end{subfigure}
\begin{subfigure}[t]{0.35\textwidth}
    \includegraphics[width=\textwidth]{bullseye}
    \caption{\label{paper1:fig:bullseye}}
  \end{subfigure}
\caption{Ventricular geometry generation. Endo and epi-cardial surfaces 
are marked on 3-D ultrasound images. Figure (\ref{paper1:fig:ultrasound_image}) shows the endocardial 
marking for a 2-D slice of one such image. Next a computational geometry is generated
from epi and endo-cardial surfaces (\ref{paper1:fig:compgeo_mesh}), and rule based
fibers are assigned (\ref{paper1:fig:compgeo_fibers}). Finally AHA segments are assigned
to the geometry (\ref{paper1:fig:compgeo_regions}), according to the standardized scheme
(\ref{paper1:fig:bullseye}).}
\end{figure}

\subsection{Parameter Estimation}
\label{paper1:sec:paramest}
Now that we have a mathematical description of cardiac motion, along
with a personalized computational geometry and target data, we next turn to the
problem of personalizing the motion model via the estimation of the
elastic parameters and the fiber contraction. As dyssynchrony is a disease 
which primarily effects the contraction properties of the ventricle,
we focus our efforts on contraction modelling and employ a very simple
personalization of stiffness properties. That is only the parameter $a$
is optimized to fit the ventricular volumes, and the other elastic parameters
are kept fixed at the values ($a_f = 1.685, b = 9.726, b_f = 15.779$),
which were obtained from a bi-axial loading experiment 
[Table 1 row 3 of \cite{Holzapfel2009}].

@@

Muscle shortening is typically present in the ventricles
throughout systole and in early diastole until the muscles fully release
their contraction. During the phase of atrial systole we do not expect
muscle contraction in the ventricle, and so we set $\gamma = 0$ for
this phase. This allows us to estimate elastic properties
independently of contraction during atrial systole, and then estimate
contraction at each point in the rest of the cardiac cycle with the
material parameters fixed. In Figure~\ref{paper1:fig:pv_loop_phases} we show
the pressure-volume loop of the patient under consideration, and
highlight the phases where we estimate the contraction and elastic
parameters.

\begin{figure}[htbp]
\centering
    \includegraphics[width=0.8\textwidth]{patient_pv_loop}
\caption{Patient pressure-volume relationship for the left
  ventricle. Measurements in the blue solid line are used to estimate
  contraction, whereas measurements in the red dashed line are used
  to estimate elasticity. }
\label{paper1:fig:pv_loop_phases}
\end{figure}

\subsection{Definition of functionals}
\label{paper1:sec:data_func}
As described in Section~\ref{paper1:sec:clinical_measurements}, the data
available for our personalization of the wall motion model are
pressure, volume and strain measurements throughout the cardiac
cycle. The pressure measurements are included in the model as a boundary condition 
via the virtual work~\eqref{paper1:eq:work-balance},
and thus our data assimilation only needs to fit the model to
the volume and strain measurements. This requires that
we define a suitable set of functionals that quantify the
model-strain and model-volume mismatches. The personalization of the wall motion model can
then be achieved by optimizing the contraction and elastic parameters
in order to minimize the total model-data mismatch.

Let $i$ denote the index of an observed cavity volume $V^i$, or strain
$\varepsilon^i$, in the cardiac cycle. Furthermore let $j \in
\{1,..,17\}$ be the index of an AHA segment $\Omega_j$, and $k \in
\{c,r,l\}$ indicate a direction: circumferential, radial or
longitudinal, respectively. Given a measurement point $i$, we define
the model-strain mismatch
\begin{equation}
 \Istrain^i =  \sum_{j= 1}^{17} \sum_{k \in \{c,r,l\}}  \left( \varepsilon_{kj}^i -  \tilde{\varepsilon}_{kj}^i \right)^2,
\label{paper1:eq:strain_functional}
 \end{equation}
for model strain $\tilde{\varepsilon}_{kj}^i$ and measured strain
$\varepsilon_{kj}^i$. The speckle tracking software we use provides the measured strain $\varepsilon_{kj}^i$. 
This strain is regionally averaged and of Lagrangian type. In order to mimic this in our model we 
define the model strain as
\begin{equation}
\varepsilon_{kj} =  \frac{1}{|\Omega_j|}\int_{\Omega_j}  \mathbf{e}_k^T \nabla \uvec \, \mathbf{e}_k \ dx,
\label{paper1:eq:g_futhese nctional}
\end{equation}
where $\ek$ denotes a unit direction field and
$|\Omega_j|$ the volume of segment $j$.

Furthermore we also define the model-volume mismatch
\begin{equation}
  \Ivol^i  =  \left( \frac{V^i - \tilde{V}^i}{V^i} \right)^2,
\end{equation}
where the model volume is calculated by the formula
\begin{equation}
  \tilde{V}^i = -\frac{1}{3} \int_{\lvendo} (\Xvec + \uvec) \cdotp J \F^{-T} \Nvec  \ dS.
  \label{paper1:eq:vol_functional}
 \end{equation}
We note that this method of calculating the model volume depends upon
$(\Xvec + \uvec)\cdotp \Nvec = 0$ at the basal plane. These conditions hold in 
our model as the basal plane is defined with 0 longitudinal coordinate and 
longitudinal displacements are also set to 0 in this plane.

In order to have a single optimization target to describe the fit to data 
we combine the strain \eqref{paper1:eq:strain_functional} and volume
\eqref{paper1:eq:vol_functional} mismatches into one single functional
\begin{equation}
  \Idata^i(\alpha) = \alpha \Ivol^i + (1 - \alpha) \Istrain^i.
\label{paper1:eq:total_functional}
\end{equation}
Here the parameter $\alpha$ controls the relative emphasis of the parameter
estimation on volume or strain matching.

In our study we consider a high dimensional representation of $\gamma$ in order
to more accurately capture the details of a dyssynchronous contraction. However this can easily lead to
an over-parametrized problem in which many parameter combinations produce the same
functional values. In order to further constrain the optimization we introduce
a 1st order Tikohonov regularization functional
\begin{equation}
 \Ismooth^i(\lambda) = \lambda \| \nabla \gamma^i \|^2, 
 \label{paper1:eq:I_smooth}
\end{equation}
where $\|\cdotp \|$ represents the standard $L^2$ norm.
This functional discriminates between $\gamma$ parameter sets based on their smoothness.
The parameter $\lambda$ can be adjusted to control the size of the functional and hence
the relative emphasis on smoothing.

\subsection{Parameter estimation as an optimization problem}
\label{paper1:sec:param_est}
The elastic parameters of the reduced
Holzapfel-Ogden law \eqref{paper1:eq:hoa} represent the passive elastic
properties of the myocardium. We personalize these properties by
adjusting the parameter $a$ to match measured left ventricular volumes.
Mathematically this problem is formulated as
\begin{equation}
\begin{aligned}
\label{paper1:eq:opt_matparam}
& \underset{a}{\text{minimize}}
& &  \sum_{i = 1}^{\Nas} \Ivol^i \\
& \text{subject to}
& & \delta W(\plv^i, a) = 0 \quad \forall \,  i \ \in \{1,...N_{AS} \},
\end{aligned}
\end{equation}
where $\delta W$ is given by \eqref{paper1:eq:workdef}, and $\Nas =
3$ indicates the total number of measurements available in atrial
systole.
